---
output: 
  pdf_document:
    latex_engine: xelatex
header-includes:
- \usepackage[fontsize=12pt]{scrextend}
editor_options: 
  markdown: 
    wrap: 72
mainfont: "Times New Roman"
---
\setcounter{page}{14}

# *7.0 Appendix*
```{r setup, results = 'hide', warning=FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
stroke = read_csv("/Users/yunkyaw/Documents/GitHub/SS-9155/healthcare-dataset-stroke-data.csv")

library(dplyr)
library(tidyr)
library(ggplot2)
library(reshape2)
library(corrplot)
library(caret) # confusion matrix
library(performanceEstimation) # SMOTE
library(car) # VIF
library(doParallel) # parallel computing
library(glmnet) # logistic regression
library(randomForest)
library(ResourceSelection) # hoslem.test
library(pROC)
library(xgboost)
```

# *7.1 EDA*

## Data Cleaning

```{r, message = FALSE, warning = FALSE}
# gender, hypertension, heart_disease, ever_married, work_type, Residence_type,
# smoking_status, target should all be categorical
stroke$gender = as.factor(stroke$gender)
stroke$hypertension = as.factor(stroke$hypertension)
stroke$heart_disease = as.factor(stroke$heart_disease)
stroke$ever_married = as.factor(stroke$ever_married)
stroke$work_type = as.factor(stroke$work_type)
stroke$Residence_type = as.factor(stroke$Residence_type)
stroke$smoking_status = as.factor(stroke$smoking_status)
# renaming response variable to targetto remove confusion
stroke$target = as.factor(stroke$stroke) 

# bmi should be numeric
stroke$bmi = as.numeric(stroke$bmi)

# removing id and stroke variables
stroke = stroke[, -c(1, 12)]

str(stroke)

# determining if there are any missing values and where
colSums(is.na(stroke))

# factor columns
factor_cols = sapply(stroke, is.factor)

# numeric columns
num_cols = sapply(stroke, is.numeric)
```

## Looking at Distribution of Numeric Variables
```{r, message=FALSE, warning=FALSE}
# convert stroke to long column dataset
melt.stroke = melt(stroke[num_cols])

# boxplot of numeric variables
ggplot(data = melt.stroke, aes(x = variable, y = value)) + 
  geom_boxplot() +
  labs(title = "Boxplots of Numeric Variables", x = "Variable", y = "Value")

# distribution of numeric variables
ggplot(data = melt.stroke, aes(x = value)) +
  stat_density(aes(y = ..scaled..)) + 
  facet_wrap(~variable, scales = "free")
```

## Looking at Distribution of Categorical Variables
```{r, warning = FALSE}
# convert to long column dataset
stroke_fact = stroke[factor_cols] %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Category")

# plotting proportion
ggplot(stroke_fact, aes(x = Category, fill = Category)) +
  geom_bar(aes(y = after_stat(prop), group = 1)) +
  facet_wrap(~ Variable, scales = "free") +
  labs(title = "Distribution of Categorical Variables", y = "Proportion") +
  theme(legend.position = "none") +
  scale_x_discrete(guide = guide_axis(angle = 45))

```

## Imputing BMI (Missing Values)
```{r}
# impute missing values with the median value
stroke$bmi[is.na(stroke$bmi)] = median(stroke$bmi, na.rm = TRUE)
summary(stroke$bmi)
```

## Correlation Analysis
```{r}
res = cor(stroke[num_cols])
corrplot(res, type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45)
```

## VIF - Looking for Multicollinearity
```{r}
# logistic regression for VIF
log_model = glm(target ~ ., family = "binomial", data = stroke)
# VIF analysis
vif_values = vif(log_model)
vif_values
```

## SMOTE for Class Imbalance
```{r}
set.seed(123)

# generating smote data
stroke_smote = smote(target ~., stroke)

table(stroke_smote$target)
```

# *7.2 Data Analysis*
## Test-Train Partition
```{r}
# 70% of data for training
smp_size = floor(0.7*nrow(stroke_smote))

set.seed(123)
# selecting which rows are for training
train_ind = sample(seq_len(nrow(stroke_smote)), size = smp_size)
train = stroke_smote[train_ind, ]  # training set
test = stroke_smote[-train_ind, ]  # testing set
```

## Logistic Regression
### Preprocessing
```{r}
# one hot encoding data - for logistic regression and xgb
one_train = model.matrix(~ . - 1, data = train)

one_test = model.matrix(~ . - 1, data = test)

# scaling avg_glucose_level and BMI - for logistic regression
log_train = as.data.frame(one_train)
log_train$avg_glucose_level = scale(log_train$avg_glucose_level)
log_train$bmi = scale(log_train$bmi)

log_test = as.data.frame(one_test)
log_test$avg_glucose_level = scale(log_test$avg_glucose_level)
log_test$bmi = scale(log_test$bmi)
```

```{r}
registerDoParallel(cores = 8) # parallel computing 

# find best lambda to apply a penalty to the features
set.seed(123)
cvfit = cv.glmnet(as.matrix(log_train[, -18]), as.matrix(log_train[, 18]),
                  nfolds = 10, alpha = 1, family = "binomial", parallel = TRUE)

# final model - logistic regression with lasso penalty
# selecting the lambda that produces the minimum MSE
log_fit = glmnet(as.matrix(log_train[, -18]), as.matrix(log_train[, 18]),
                 alpha = 1, family = "binomial", lambda = cvfit$lambda.min)

stopImplicitCluster() # stop parallel computing

# coefficients
round(coef(log_fit), 3)

# for log-odds calculation
# exp(coef(log_fit))
```

### Goodness of Fit
```{r}
# producing probability of each row belonging to a class
log_prob = predict(log_fit, newx = as.matrix(log_test[, -18]), type = "response")

hoslem.test(as.numeric(log_test[, 18]), log_prob, g = 10)
# good fit as p > 0.05
```

### Generating Predictions and Confusion Matrix
```{r}
# predicting the class of each row with a 0.50 threshold
log_pred = ifelse(log_prob > 0.5, 1, 0)

# confusion matrix
confusionMatrix(as.factor(log_pred), as.factor(log_test[, 18]))
```

### ROC Analysis
```{r, warning = FALSE, message = FALSE}
# developing and plotting ROC curve
roc_obj = roc(response = test$target, predictor = log_prob)

# calculating area under the curve
AUC = auc(roc_obj)

# getting threshold, sensitivity, accuracy, and AUC for current model
roc_log = c(coords(roc_obj, "b", ret = c("threshold", "se", "sp", "accuracy"), 
                   best.method = "youden"), AUC)

names(roc_log) = c("Threshold", "Sensitivity", "Specificity", "Accuracy", "AUC")

t(roc_log)
```

### Changing Threshold
```{r}
log_pred_adj = ifelse(log_prob > 0.3990413, 1, 0)

# confusion matrix
confusionMatrix(as.factor(log_pred_adj), as.factor(log_test[, 18]))
```

## Random Forest
```{r}
registerDoParallel(cores = 8) # parallel computing 

set.seed(123)

# control function for grid search, setting 10-fold cv repeated 3 times
control = trainControl(method = "repeatedcv", number = 10, repeats = 3,
                       search = "grid")

# defining the grid parameters - mtry (number of splits per tree)
grid = expand.grid(mtry = c(1:10))

# training the random forest model with 10-fold cross validation and 1000 trees
rf_gridsearch = train(target~., data=train, method="rf", tuneGrid = grid,
                      trControl = control)

plot(rf_gridsearch)
# determine optimal mtry = 5

# optimal random forest 
rf_gridsearch
```

### Generating Optimal Model
```{r}
rfModel = randomForest(target ~ ., data = train, proximity = T, ntree = 1000,
                       mtry = 5) 
stopImplicitCluster() # stop parallel computing
rfModel
```

### Visualize Important Features
```{r}
varImpPlot(rfModel)
```

### Generating Predictions and Confusion Matrix
```{r}
# probabilistic predictions
rf_prob = predict(rfModel, newdata = test[, -11], type = "prob")

# class predictions
rf_pred = predict(rfModel, newdata = test[, -11], type = "response")

confusionMatrix(rf_pred, test$target)
```

## XGBoost
### Preprocessing 
```{r}
# use one-hot encoded variables, creating DMatrix data structure for xgb
# data = independent variables, label = dependent variable
xgb_train = xgb.DMatrix(data = as.matrix(one_train[, 1:17]),
                        label = as.numeric(one_train[, 18]))
xgb_test = xgb.DMatrix(data = as.matrix(one_test[, 1:17]),
                       label = as.numeric(one_test[, 18]))

# set parameters
params = list(
  booster = "gbtree",            # decision tree
  eta = 0.1,                     # learning rate
  max_depth = 5,                 # maximum tree depth
  objective = "multi:softprob",  # generate probabilities for classes
  nthread = 10,                  # nos of CPU for model use
  num_class = 2                  # nos of classes of target
)
```

### Cross-Validation to Find Optimal Model
```{r, result = FALSE, message = FALSE}
# utilizing cross-validation to determine the optimal number of rounds
set.seed(123)
cv_xgb = xgb.cv(params = params, data = xgb_train, nrounds=2000, nfold = 10,
                verbose = 0, early_stopping_rounds = 10)

# find optimal nrounds = 110
# train-mlogloss:0.173914+0.004857	test-mlogloss:0.405670+0.034391
```

### Generating Optimal Model 
```{r, result = 'hide'}
set.seed(123)
xgb_model = xgboost(params = params, data = xgb_train, nrounds = 110, verbose = 0)
# train-mlogloss:0.180190 
```

### Generating Predictions and Confusion Matrix
```{r}
# probabilistic predictions
xgb_prob = predict(xgb_model, xgb_test, reshape = TRUE)

colnames(xgb_prob) = levels(train$target)

# generate class predictions for probability of being positive
xgb_pred = ifelse(xgb_prob[, 2] > 0.5, 1, 0)

confusionMatrix(as.factor(xgb_pred), test$target)
```

### Important Features
```{r}
xgb_feat = xgb.importance(feature_names = colnames(xgb_train), model = xgb_model)
xgb.plot.importance(xgb_feat)
```

# *7.3 Comparison*
## ROC Curves
```{r, warning = FALSE, message=FALSE}
# logistic regression
roc_log = roc(response = test$target, predictor = log_prob)
# AUC = 0.8357

# random forest - looking at the probability of a positive observation
roc_rf = roc(response = test$target, predictor = rf_prob[, "1"])
# AUC = 0.8997

# xgboost - looking at the probability of a positive observation
roc_xgb = roc(response = test$target, predictor = xgb_prob[, 2])
# AUC = 0.8791

# plotting
plot(roc_log, legacy.axes = FALSE, cex.lab = 2, col = "green")
lines(roc_log, col = "green")
lines(roc_rf, col = "red")
lines(roc_xgb, col = "blue")
legend("bottomright", legend = c("Logistic Regression (AUC = 0.8357)",
                                 "Random Forest (AUC = 0.8997)",
                                 "XGBoost (AUC = 0.8791)"), 
       col = c("green", "red", "blue"), lty = 1)
```

## Table of Metrics
```{r}
# retrieved metric values from confusionMatrix for each model
metrics = c("Accuracy", "Sensitivity", "Specificity", "AUC")
log_metr = c(0.7667, 0.7, 0.8565, 0.8357)
rf_metr = c(0.826, 0.81, 0.8475, 0.8997)
xgb_metr = c(0.782, 0.77, 0.7982, 0.8791)

results_table = data.frame(
  Metric = metrics,
  Logistic_Regression = log_metr,
  Random_Forest = rf_metr,
  XGBoost = xgb_metr
)

t(results_table)
```
